Using functor categories to generate intermediate code with Agda

### 2025-04-23 draft.tex
introduction and preparation almost complete

\documentclass[12pt,twoside,a4paper]{report}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{graphicx}
\usepackage{fancyhdr}
\usepackage{enumerate}
\usepackage[vmargin=2.5cm, hmargin=2.5cm]{geometry}
\usepackage{hyperref}
\usepackage{fontspec}
\usepackage{unicode-math}
\usepackage{stackengine}
\usepackage{minitoc}
\usepackage{amsthm}
\usepackage{listings}

\setmainfont{Linux Libertine O}
\setsansfont{Linux Biolinum O}

\graphicspath{{./fig/category_theory/}}

@@

@@

@@

    The grammer of the target language is specified in Chapter 3.

@@

    \section{Motivation and related work}
        The denotational semantics of Algol-like languages can be structured as a presheaf category over stack descriptors, which has been shown by Reynolds \cite{essence} and Oles \cite{Oles_1} \cite{Oles_2}. By interpreting the source language into this category, where objects of the category represent instruction sequences parameterised by stack layouts, the semantic model directly yields a compiler. The mathematical structure of the compiler has been specified by Reynolds in his paper ``Using Functor Categories to Generate Intermediate Code'' \cite{Reynolds}.

@@

        (Example of Agda)

        Dependently typed languages provide a natural framework for expressing functor categories is proven both theoretically and practically. There have been dependent-type-theoretic model of categories \cite{Dybjer}, and it has been shown that functor categories arise naturally as dependent function types \cite{Jacobs}. A formalisation of Category Theory, including Cartesian Closed Categories, functors and presheaves has been developed in Agda by Hu and Caratte \cite{Cat_Agda}. Other proof assistants, such as Isabelle/Hol, does not have a dependently typed language structure, and thus cannot express the functor categories as naturally as Agda.

@@

    \section{Contributions}
        Addressed the two motivations presented in 1.1 and contributed to the following:
        \begin{quote}
            \savedmotivationI
            We implemented the compiler from the source language to the target language in Agda.

@@

    \section{Category theory background}
        Category 
        \subsection{Category}
        \begin{definition*}[Category]
            A \emph{category} $\mathcal{C}$ is specified by the following:
            \begin{itemize}
                \item 
                    a collection of objects $\textbf{obj}(\mathcal{C})$, whose elements are called $\mathcal{C}$-objects;
                \item 
                    for each $X, Y \in \textbf{obj}(\mathcal{C})$, a collection of morphisms $\mathcal{C}{(X,Y)}$, whose elements are called $\mathcal{C}$-morphisms from $X$ to $Y$;
                \item 
                    for each $X \in \textbf{obj}(\mathcal{C})$, an element $\textbf{id}_X \in \mathcal{C}{(X,X)}$ called the identity morphism on $X$;
                \item 
                    for each $X, Y, Z \in \textbf{obj}(\mathcal{C})$, a function 
                    \[\begin{aligned}
                        \mathcal{C}{(X,Y)} \times \mathcal{C}{(Y,Z)} &\to \mathcal{C}{(X,Z)} \\
                        (f,g) &\mapsto g \circ f
                    \end{aligned}\]
                    called the composition of morphisms;
            \end{itemize}
            satisfying the following properties:
            \begin{itemize}
                \item 
                    \textbf{(Unit Laws)}
                    for all $X, Y \in \textbf{obj}(\mathcal{C})$ and $f \in \mathcal{C}{(X,Y)}$, we have:
                    \begin{equation} \label{law: unit}
                        \textbf{id}_Y \circ f = f = f \circ \textbf{id}_X
                    \end{equation}
                \item
                    \textbf{(Associativity Law)}
                    for all $X, Y, Z, W \in \textbf{obj}(\mathcal{C})$ and $f \in \mathcal{C}{(X,Y)}$, $g \in \mathcal{C}{(Y,Z)}$, $h \in \mathcal{C}{(Z,W)}$, we have:
                    \begin{equation} \label{law: associativity}
                        h \circ (g \circ f) = (h \circ g) \circ f
                    \end{equation}
            \end{itemize}

            [Considering adding different notations for morphisms, composition, identity morphisms, etc.]

            [Considering adding a diagram for the definition of category]

            [Considering adding examples of categories]

@@

        \end{definition*}
        [Considering adding examples of isomorphism]

@@

                \item 
                    for each $X, Y \in \textbf{obj}(\mathcal{C})$, a function 
                    \[\begin{aligned}
                        \mathcal{C}{(X,Y)} &\to \mathcal{D}{(F(X),F(Y))} \\
                        f &\mapsto F(f)
                    \end{aligned}\]
            \end{itemize}
            satisfying the following properties:
            \begin{itemize}
                \item 
                    for all $X, Y \in \textbf{obj}(\mathcal{C})$ and $f \in \mathcal{C}{(X,Y)}$, we have:
                    \begin{equation} \label{law: functor_id}
                        F(\textbf{id}_X) = \textbf{id}_{F(X)}
                    \end{equation}
                \item
                    for all $X, Y, Z \in \textbf{obj}(\mathcal{C})$ and $f \in \mathcal{C}{(X,Y)}$, $g \in \mathcal{C}{(Y,Z)}$, we have:
                    \begin{equation} \label{law: functor_comp}
                        F(g \circ f) = F(g) \circ F(f)
                    \end{equation}
            \end{itemize}
        \end{definition*}
        [Considering adding examples of functors: e.g. free functor, forgetful functor, etc.]

@@

        \end{definition*}
        [Considering adding examples of natural transformations]

@@

        \subsection{Yoneda lemma}
        \begin{definition*}[Yoneda functor]
            Given a category $\mathcal{C}$, the \emph{Yoneda functor} $y: \mathcal{C} \to \mathcal{C}^{\mathcal{C}^{op}}$ is defined as follows:
            \begin{itemize}
                \item 
                    for each $X \in \textbf{obj}(\mathcal{C})$, $y(X)$ is the functor $y(X): \mathcal{C}^{op} \to \textbf{Set}$ defined as:
                    \[ y(X)(Y) = \mathcal{C}{(Y,X)} \]
                    for all $Y \in \textbf{obj}(\mathcal{C})$;
                \item
                    for each $X, Y \in \textbf{obj}(\mathcal{C})$ and $f \in \mathcal{C}{(X,Y)}$, $y(f)$ is the natural transformation $y(f): y(X) \to y(Y)$ defined as:
                    \[ y(f)_Y = f \circ - : y(X)(Y) \to y(Y)(Y) \]
                    for all $Y \in \textbf{obj}(\mathcal{C})$.
                    In other words, $y(f)_Y$ is the function that takes an element $g \in y(X)(Y) = \mathcal{C}{(Y,X)}$ and returns the element $f \circ g \in y(Y)(Y) = \mathcal{C}{(Y,Y)}$.
                    The function $y(f)_Y$ is a morphism in $\mathcal{C}^{\mathcal{C}^{op}}$ from $y(X)(Y)$ to $y(Y)(Y)$.
                    In other words, $y(f)_Y$ is a morphism in $\mathcal{C}^{\mathcal{C}^{op}}$ from the functor $y(X)$ to the functor $y(Y)$.
                    The morphism $y(f)_Y$ is defined for all $Y \in \textbf{obj}(\mathcal{C})$.
                    In other words, $y(f)$ is a natural transformation from the functor $y(X)$ to the functor $y(Y)$.
            \end{itemize}
        \end{definition*}

        \begin{definition*}[Yoneda lemma]
            Given a category $\mathcal{C}$, the \emph{Yoneda lemma} states that for any functor $F: \mathcal{C} \to \textbf{Set}$, there is a natural isomorphism:
            \[ \begin{aligned}
                \textbf{Nat}(y(X),F) &\cong F(X) \\
                \theta &\mapsto \theta_X
            \end{aligned} \]
            for all $X \in \textbf{obj}(\mathcal{C})$, where $\textbf{Nat}(y(X),F)$ is the set of natural transformations from the functor $y(X)$ to the functor $F$.
        \end{definition*}

        \subsection{Presheaf}
        \begin{definition*}[Presheaf]
            Given a category $\mathcal{C}$, a \emph{presheaf} on $\mathcal{C}$ is a functor $F: \mathcal{C}^{op} \to \textbf{Set}$.
            A presheaf is a contravariant functor, which means that it reverses the direction of morphisms.
            In other words, a presheaf is a functor that takes objects in $\mathcal{C}$ and assigns them sets, and takes morphisms in $\mathcal{C}$ and assigns them functions between the corresponding sets.
            The presheaf $F$ is defined as follows:
            \begin{itemize}
                \item 
                    for each $X \in \textbf{obj}(\mathcal{C})$, $F(X)$ is a set;
                \item
                    for each $X, Y \in \textbf{obj}(\mathcal{C})$ and $f \in \mathcal{C}{(X,Y)}$, $F(f)$ is a function $F(Y) \to F(X)$.
                    In other words, $F(f)$ is a function that takes an element $g \in F(Y)$ and returns the element $f \circ g \in F(X)$.
                    The function $F(f)$ is a morphism in $\mathcal{C}^{op}$ from $F(Y)$ to $F(X)$.
                    In other words, $F(f)$ is a morphism in $\mathcal{C}^{op}$ from the functor $F(Y)$ to the functor $F(X)$.
                    The morphism $F(f)$ is defined for all $X, Y \in \textbf{obj}(\mathcal{C})$ and $f \in \mathcal{C}{(X,Y)}$.
                    In other words, $F(f)$ is a natural transformation from the functor $F(Y)$ to the functor $F(X)$.
            \end{itemize}
        \end{definition*}

        \subsection{Cartesian closed structure in presheaf categories}

@@

        \subsection{Dependent Types}

        \subsection{Curry-Howard-Lambek correspondence}

        \subsection{Equality, congruence and substitution}

        % \subsection{Proof Example}

        \subsection{Standard library}

        \subsection{Interactive programming with holes}
        A feature of Agda is that it allows us to write programs with holes interatively. A hole is a placeholder for a term that we have not yet defined. By leaving holes in place of undefined terms, we can write programs that are incomplete but still type-check, and Agda's type checker will guide completion of the program: the context window displays inferred types of the holes, available variables and candidate terms with their types. Holes also supports case split and refinement, which means we can fill in a hole partially and split it into smaller holes.

@@

    \section{Requirement Analysis}

@@

>>>>

\documentclass[12pt,twoside,a4paper]{report}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{graphicx}
\usepackage{fancyhdr}
\usepackage{enumerate}
\usepackage[vmargin=2.5cm, hmargin=2.5cm]{geometry}
\usepackage{hyperref}
\usepackage{fontspec}
\usepackage{unicode-math}
\usepackage{stackengine}
\usepackage{minitoc}
\usepackage{amsthm}
\usepackage{listings}
\usepackage{xcolor}
\usepackage{xeCJK}

\setmainfont{Linux Libertine O}
\setsansfont{Linux Biolinum O}
\setCJKmainfont{Noto Serif CJK JP}

\graphicspath{{./fig/category_theory/}{./fig/agda/lagda/latex/}}

@@

\theoremstyle{definition}
\newtheorem*{theorem*}{Theorem}

@@

\newcommand{\secref}[1]{\S\ref{#1}}

\definecolor{mediumblue}{HTML}{0000CD}
\definecolor{green}{HTML}{008B00}
\newcommand{\mb}[1]{\textcolor{mediumblue}{#1}}
\newcommand{\gn}[1]{\textcolor{green}{#1}}

\newcommand{\bN}{\mathbb{N}}

\newcommand{\yo}{\mathord{\text{\kern0.03em \scalebox{0.8}{よ}\kern0.03em}}}

@@

    The grammer of the target language is specified in Chapter 3. [Add exact section/subsection here]

@@

    \section{Motivation and related work} \label{sec: motivation}
        The denotational semantics of Algol-like languages can be structured as a presheaf category over stack descriptors, which has been shown by Reynolds \cite{essence} and Oles \cite{Oles_1} \cite{Oles_2}. By interpreting the source language into this category, where objects of the category represent instruction sequences parameterised by stack layouts, the semantic model directly yields a compiler. The mathematical structure of the compiler has been specified by Reynolds in his paper ``Using Functor Categories to Generate Intermediate Code'' \cite{Reynolds}.

@@

        Dependently typed languages provide a natural framework for expressing functor categories is proven both theoretically and practically. There have been dependent-type-theoretic model of categories \cite{Dybjer}, and it has been shown that functor categories arise naturally as dependent function types \cite{Jacobs}. A formalisation of Category Theory, including cartesian closed categories, functors and presheaves has been developed in Agda by Hu and Caratte \cite{Cat_Agda}. Other proof assistants, such as Isabelle/Hol, does not have a dependently typed language structure, and thus cannot express the functor categories as naturally as Agda. Agda also provides an interactive environment for writing and verifying programs, which will be further discussed in \secref{subsec: holes}.

@@

    \section{Contributions}
        Addressed the three motivations presented in \secref{sec: motivation} and contributed to the following:
        \begin{quote}
            \savedmotivationI
            We implemented the compiler from the source language to the target language in Agda.

@@

    \section{Category theory background} \label{sec: cat}
        Category theory provides a high-level abstraction from which we can reason about the structure of mathematical objects and their relationships. It provides us a ``bird's eye view'' of the mathematics which enable us to spot patterns that are difficult to see in the details. \cite{basic_cat} More specifically, it provides a ``purer'' view of functions that is not derived from sets. \cite{scott-lambda} Compared to set theory which is ``element-oriented'', category theory is ``function-oriented'' and ``morphism-oriented''. We understand structures not via elements but by how they transform into each other.

        Notation in category theory is similar to that in set theory and type theory. For example, $A, B, \dots$ are used to denote objects, sets, or types, and arrows are used to denote morphisms or functions (e.g.\ $A \to B$). There is a deeper connection between category theory and type theory. Curry-Howard correspondence \cite{curry-howard} states that there is a correspondence between logic and type theory, where propositions correspond to types and proofs correspond to terms. This correspondence was later extended by Joachim Lambek, who showed that cartesian closed categories provide a natural semantic setting for the simply typed lambda calculus. \cite{lambek} Under this perspective, types become objects in a category, and programs correspond to morphisms. We will further explore this connection in \secref{subsec: curry-howard-lambek}.

        The following is a brief introduction to the basic concepts of category theory, which is based on the work of Leinster \cite{basic_cat}, 
        Riehl \cite{cat_context}, and the lecture notes of Part II Category Theory by Andrew Pitts and Marcelo Fiore \cite{cat_lecture_notes}.

        \subsection{Category}
        \begin{definition*}[Category]
            A \emph{category} $\mathcal{C}$ is specified by the following:
            \begin{itemize}
                \item 
                    a collection of objects $\textbf{obj}(\mathcal{C})$, whose elements are called $\mathcal{C}$-objects;
                \item 
                    for each $X, Y \in \textbf{obj}(\mathcal{C})$, a collection of morphisms $\mathcal{C}{(X,Y)}$, whose elements are called $\mathcal{C}$-morphisms from $X$ to $Y$ (e.g.\ $f : X \to Y$);
                \item 
                    for each $X \in \textbf{obj}(\mathcal{C})$, an element $\textbf{id}_X \in \mathcal{C}{(X,X)}$ called the identity morphism on $X$;
                \item 
                    for each $X, Y, Z \in \textbf{obj}(\mathcal{C})$, a function 
                    \[\begin{aligned}
                        \mathcal{C}{(X,Y)} \times \mathcal{C}{(Y,Z)} &\to \mathcal{C}{(X,Z)} \\
                        (f,g) &\mapsto g \circ f
                    \end{aligned}\]
                    called the composition of morphisms;
            \end{itemize}
            satisfying the following properties:
            \begin{itemize}
                \item 
                    \textbf{(Unit Laws)}
                    for all $X, Y \in \textbf{obj}(\mathcal{C})$ and $f \in \mathcal{C}{(X,Y)}$, we have:
                    \begin{equation} \label{law: unit}
                        \textbf{id}_Y \circ f = f = f \circ \textbf{id}_X
                    \end{equation}
                \item
                    \textbf{(Associativity Law)}
                    for all $X, Y, Z, W \in \textbf{obj}(\mathcal{C})$ and $f \in \mathcal{C}{(X,Y)}$, $g \in \mathcal{C}{(Y,Z)}$, $h \in \mathcal{C}{(Z,W)}$, we have:
                    \begin{equation} \label{law: associativity}
                        h \circ (g \circ f) = (h \circ g) \circ f
                    \end{equation}
            \end{itemize}
        \end{definition*}

        An example of category is the category of sets, denoted as $\mathcal{Set}$, specifed by:
        \begin{itemize}
            \item 
                the objects of $\mathcal{Set}$ are a fixed universe of sets;
            \item
                for each $X, Y \in \textbf{obj}(\mathcal{Set})$, the morphisms from $X$ to $Y$ in $\mathcal{Set}$ are the functions from $X$ to $Y$;
            \item
                the identity morphism on $X$ in $\mathcal{Set}$ is the identity function on $X$;
            \item
                the composition of morphisms in $\mathcal{Set}$ is defined as the composition of functions.
        \end{itemize}
        Associativity law and unit laws are satisfied in $\mathcal{Set}$, since the composition of functions is associative and the identity function compose with any function is the function itself.

@@

        [Considering adding examples of isomorphism]

@@

                \item 
                    for each $X, Y \in \textbf{obj}(\mathcal{C})$, a function 
                    \[\begin{aligned}
                        \mathcal{C}{(X,Y)} &\to \mathcal{D}{(F(X),F(Y))} \\
                        f &\mapsto F(f)
                    \end{aligned}\]
            \end{itemize}
            satisfying the following properties:
            \begin{itemize}
                \item 
                    for all $X, Y \in \textbf{obj}(\mathcal{C})$ and $f \in \mathcal{C}{(X,Y)}$, we have:
                    \begin{equation} \label{law: functor_id}
                        F(\textbf{id}_X) = \textbf{id}_{F(X)}
                    \end{equation}
                \item
                    for all $X, Y, Z \in \textbf{obj}(\mathcal{C})$ and $f \in \mathcal{C}{(X,Y)}$, $g \in \mathcal{C}{(Y,Z)}$, we have:
                    \begin{equation} \label{law: functor_comp}
                        F(g \circ f) = F(g) \circ F(f)
                    \end{equation}
            \end{itemize}
        \end{definition*}

        [Considering adding examples of functors: e.g. free functor, forgetful functor, etc.]

@@

        [Considering adding examples of natural transformations]

@@

        \subsection{Presheaf category}
        \begin{definition*}[Presheaf]
            Given a category $\mathcal{C}$, a \emph{presheaf} on $\mathcal{C}$ is a functor $F: \mathcal{C}^{\textbf{op}} \to \textbf{Set}$.
            A presheaf is a contravariant functor, which means that it reverses the direction of morphisms.
            In other words, a presheaf is a functor that takes objects in $\mathcal{C}$ and assigns them sets, and takes morphisms in $\mathcal{C}$ and assigns them functions between the corresponding sets.
            The presheaf $F$ is defined as follows:
            \begin{itemize}
                \item 
                    for each $X \in \textbf{obj}(\mathcal{C})$, $F(X)$ is a set;
                \item
                    for each $X, Y \in \textbf{obj}(\mathcal{C})$ and $f \in \mathcal{C}{(X,Y)}$, $F(f)$ is a function $F(Y) \to F(X)$.
            \end{itemize}
        \end{definition*}

        \begin{definition*}[Presheaf category]
            Given a category $\mathcal{C}$, the \emph{presheaf category} $\mathcal{P}(\mathcal{C})$ is the functor category $\mathcal{Set}^{\mathcal{C}^{\textbf{op}}}$.
            \begin{itemize}
                \item 
                    the objects of $\mathcal{P}(\mathcal{C})$ are all presheaves on $\mathcal{C}$;
                \item
                    given two presheaves $F,G: \mathcal{C}^{\textbf{op}} \to \mathcal{Set}$, the morphisms from $F$ to $G$ in $\mathcal{P}(\mathcal{C})$ are all natural transformations $\theta : F \to G$.
            \end{itemize}
        \end{definition*}

        \subsection{Yoneda lemma}
        \begin{definition*}[Yoneda functor]
            Given a category $\mathcal{C}$, the \emph{Yoneda functor} $\yo: \mathcal{C} \to \mathcal{Set}^{\mathcal{C}^{\textbf{op}}}$ is defined as follows:
            \begin{itemize}
                \item 
                    for each $X \in \textbf{obj}(\mathcal{C})$, $\yo(X)$ is the functor $y(X) : \mathcal{C}^{\textbf{op}} \to \mathcal{Set}$ defined as:
                    \begin{equation} \label{eq: yoneda_element}
                        \yo(X)(Y) = \mathcal{C}{(Y,X)}
                    \end{equation}
                    for all $Y \in \textbf{obj}(\mathcal{C})$;
                \item
                    for each $X, Y \in \textbf{obj}(\mathcal{C})$ and $f \in \mathcal{C}{(X,Y)}$, $\yo(f)$ is the morphism $\yo(X) \to \yo(Y)$ defined a natural transformation whose component at any given $Z \in \mathcal{C}^{\textbf{op}}$ is given by:
                    \begin{equation} \label{eq: yoneda_morphism}
                        \begin{aligned}
                            (\yo(f))_Z : \mathcal{C}{(Z,Y)} &\to \mathcal{C}{(Z,X)} \\
                            g &\mapsto g \circ f
                        \end{aligned}
                    \end{equation}
                    for all $Z \in \textbf{obj}(\mathcal{C})$.
                \end{itemize}

        \begin{theorem*}[Yoneda lemma]
            Given a category $\mathcal{C}$, the \emph{Yoneda lemma} states that for each object $X \in \textbf{obj}(\mathcal{C})$ and any functor $F: \mathcal{C} \to \textbf{Set}$, there is a natural isomorphism:
            \begin{equation} \label{eq: yoneda_lemma}
                \mathcal{Set}^{\mathcal{C}^{\textbf{op}}}(\yo(X), F) \cong F(X)
            \end{equation}
        \end{theorem*}

        \subsection{Cartesian closed structure in presheaf categories}
        \begin{theorem*}[Cartesian closed structure in presheaf categories]
            Given a small category $\mathcal{C}$, the presheaf category $\mathcal{Set}^{\mathcal{C}^{\textbf{op}}}$ is a Cartesian closed category.
            \begin{itemize}
                \item 
                    The terminal object in $\mathcal{Set}^{\mathcal{C}^{\textbf{op}}}$ is the constant functor $\mathbf{1} : \mathcal{C}^{\textbf{op}} \to \mathcal{Set}$, given by 
                    \begin{equation}
                        \begin{cases}
                            \mathbf{1}(X) = \{\ast\} & \text{for all } X \in \textbf{obj}(\mathcal{C}) \\
                            \mathbf{1}(f) = \textbf{id}_{\{\ast\}} & \text{for all } f \in \mathcal{C}{(X,Y)}
                        \end{cases}
                    \end{equation}

                \item
                    The binary product in $\mathcal{Set}^{\mathcal{C}^{\textbf{op}}}$ is given by the product of functors, which is defined as follows:
                    \begin{equation}
                        \begin{aligned}
                            (F \times G)(X) &= F(X) \times G(X) \\
                            (F \times G)(f) &= F(f) \times G(f)
                        \end{aligned}
                    \end{equation}

                \item
                    The exponential in $\mathcal{Set}^{\mathcal{C}^{\textbf{op}}}$ is given by the Yoneda lemma,
                    \begin{equation}
                        \begin{aligned}
                            G^{F}(X) &= \mathcal{Set}^{\mathcal{C}^{\textbf{op}}}(\yo(X) \times F, G) \\
                            G^{F}(f)(\theta) &= \theta \circ (\yo(f) \times \textbf{id}_F) \quad \text{for all } \theta \in \mathcal{Set}^{\mathcal{C}^{\textbf{op}}}(\yo(Y) \times F, G)
                        \end{aligned}
                    \end{equation}
            \end{itemize}
        \end{theorem*}

@@

        \[\includegraphics[width=\textwidth]{example.pdf}\]

        Here we define a datatype $\mb{\bN}$ for natural numebers. $\mb{\bN}$ itself has a type $\mb{\mathsf{Set}}$, which is the type of all small types. The natural number is defined as a recursive datatype with two constructors: $\gn{\mathsf{zero}}$ and $\gn{\mathsf{suc}}$, where 
        $\gn{\mathsf{zero}}$ is the base case and $\gn{\mathsf{suc}}$ is the inductive case. The $\gn{\mathsf{zero}}$ constructor has type $\mb{\bN}$, and the $\gn{\mathsf{suc}}$ constructor has type $\mb{\bN \to \bN}$.

        In order to define the plus function $\mb{\_+\_}$, we use pattern matching to match the input argument with the constructors of the datatype. The first case is the base case, where we match the input argument with $\gn{\mathsf{zero}}$. In this case, we return the second argument. The second case is the inductive case, where we match the input argument with $\gn{\mathsf{suc}}$. In this case, we return $\gn{\mathsf{suc}}$ applied to the result of adding one to the second argument.

        \subsection{Dependent Types}
        A dependent type is a type that depends on a value. In terms of type judgement, simple types are in form of
        \[ x_1 : T_1, x_2 : T_2, \dots, x_n : T_n \vdash t(x_1, \dots, x_n) : T \]
        In contrast, dependent types are in form of
        \[ x_1 : T_1, x_2 : T_2, \dots, x_n : T_n \vdash t(x_1, \dots, x_n) : T(x_1, \dots, x_n) \]
        or more generally
        \[ x_1 : T_1, x_2 : T_2(x_1), \dots, x_n : T_n(x_1, \dots, x_{n-1}) \vdash t(x_1, \dots, x_n) : T(x_1, \dots, x_n) \]

        A classical example of dependent types in Agda is the type of vectors, which are lists with a length.

        \[\includegraphics[width=\textwidth]{vector.pdf}\]

        Here we define a datatype $\mb{\mathsf{Vec}}$ for vectors. The type of vectors is dependent on the length of the vector. With dependent types, we can encode properties directly into types and ensure that they are satisfied at compile time.

        \subsection{Curry-Howard-Lambek correspondence} \label{subsec: curry-howard-lambek}
        As introduced in \secref{sec: cat}, Curry-Howard correspondence states that there is a correspondence between logic and type theory, and Lambek extended this correspondence to show that cartesian closed categories provide a natural semantic setting for the simply typed lambda calculus. They can be summarised as follows:
        \begin{table}[htb!]
            \centering
            \begin{tabular}{|c|c|c|}
                \hline
                \textbf{Logic} & \textbf{Type theory} & \textbf{Category theory} \\
                \hline
                Proposition & Type & Object \\
                Proof & Term & Morphism \\
                Falsity & Empty type & Initial object \\
                Truth & Unit type & Terminal object \\
                Implication & Function type & Exponential object \\
                Conjunction & Product type & Product \\
                Disjunction & Sum type & Coproduct \\
                Universal quantification & Dependent product type &  \\
                Existential quantification & Dependent sum type &  \\
                \hline
            \end{tabular}
            \caption{Curry-Howard-Lambek correspondence}
            \label{tab: curry-howard-lambek}
        \end{table}

        \subsection{Equality, congruence and substitution}
        Equality here refers to the propositional equality. In Agda it is defined as
        \[\includegraphics[width=\textwidth]{equality.pdf}\]
        With equality as a type, whenever we want to prove that two terms are equal, we need to provide a witness of the equality. For example if we want to prove $x \equiv y$, we write out a term of type $\mb{\mathsf{x = y}}$, and then give its definition, which constructs a proof of the equality. A simple proof that directly uses the definition of equality is as follows:
        \[\includegraphics[width=\textwidth]{sym_trans.pdf}\]
        Here we are able to prove the symmetry and transitivity of equality by using the definition of equality. Those two properties are very useful for later proofs.

        We can also have more complex proof with congruence and substitution, which can also be directly derived from the definition of equality as follows:
        \[\includegraphics[width=\textwidth]{cong_subst.pdf}\]
        Congruence is a property of equality that states that if two terms are equal, then they can be substituted for each other in any context. Substitution is a property where we can get a new proof by replacing a term in a proof with another term that is equal to it.

        Here is a simple example of how congruence can be used in a proof:
        \[\includegraphics[width=\textwidth]{cong_example.pdf}\]

        In this example, we want to prove that $\gn{\mathsf{zero}}$ is the right identity of the plus function. We do an inductive proof by pattern matching on the first argument of the plus function.

        In the base case, we have $\gn{\mathsf{zero}} \mb{{}+{}} \gn{\mathsf{zero}} \equiv \gn{\mathsf{zero}}$, which is trivially true by the definition of the plus function (zero is defined to be the left identity of the plus function).

        In the inductive case, we need to show $\gn{\mathsf{suc}}\ n \mb{{}+{}} \gn{\mathsf{zero}} \equiv \gn{\mathsf{suc}}\ n$. We can use the definition of the plus function to rewrite the left-hand side as $\gn{\mathsf{suc}}\ (n \mb{{}+{}} \gn{\mathsf{zero}})$. By the inductive hypothesis, we know that $n \mb{{}+{}} \gn{\mathsf{zero}} \equiv n$, so we can substitute $n$ for $n \mb{{}+{}} \gn{\mathsf{zero}}$ in the right-hand side by congruence, and we are done.

        \subsection{Proof Example}
        [decide exact example here]

        \subsection{Standard library}
        The Agda standard library \cite{agda_std} is a collection of modules that provide a wide range of useful functions and types. It includes modules for basic data types, such as natural numbers, lists, and vectors. In my implementation, however, I defined most of the basic data types and functions from scratch for the following two reasons:
        \begin{itemize}
            \item
                I wanted to have a better understanding of the basic data types and functions, so I decided to implement them from scratch.
            \item
                I need a particular representation of minus operation that ensures the input is always valid. This is further explained in [to add]
        \end{itemize}

        With standard library, we can use actual numbers $1, 2, 3, \dots$ instead of calling the constructor $\gn{\mathsf{suc}}$ multiple times. This is a convenient feature for testing. Agda provides a way to link self-defined functions to the standard library with a BUILTIN programa. This allows me to use a convenient syntax for numbers, while still using the self-defined functions in the implementation.

        \subsection{Interactive programming with holes} \label{subsec: holes}
        A feature of Agda is that it allows us to write programs with holes interatively. A hole is a placeholder for a term that we have not yet defined. By leaving holes in place of undefined terms, we can write programs that are incomplete but still type-check, and Agda's type checker will guide completion of the program: the context window displays inferred types of the holes, available variables and candidate terms with their types. Holes also supports case split and refinement, which means we can fill in a hole partially and split it into smaller holes.

@@

    \section{Requirement Analysis}
    To complete a compiler in Agda we need to implement the following components:
    \begin{itemize}
        \item 
            A file source.agda that record the syntax of the source language.

        \item
            A file target.agda that record the syntax of the target language.

        \item
            A file compiler.agda that uses source.agda and target.agda as modules, and write functions whose input is a term in the source language and output is a term in the target language.
    \end{itemize}
    The success criteria of the project is that the compiler can compile a term in the source language to a term in the target language, and the output term is well-typed in the target language.

@@

    \section{Summary}
    \begin{itemize}
        \item 
        requirement Analysis

        \item
        tools and engineering approach

    \end{itemize}
